Relative Stable Maps in Genus One via Central Alignments

### 2019-01-02 GenusOneRelativeMaps.tex
Wrote some more in Section 4. The amount of words is getting a bit mental, but with some massaging it will end up being more readable.

\noindent

We now focus our attention on the logarithmic divisors, providing a recursive description for these loci: that is, a description in terms of fibre products of smaller moduli spaces. This is the technically most difficult part of the paper. The recursive description bears some similarities with the recursive desription of the boundary of the absolute space $\VZ_{1,n}(\PP^N,d)$ given in \cite{VZ}; in keeping with the theme of this paper, however, we will see that the logarithmic structures provide a convenient framework for dealing with the combinatorics.

The end result is that we obtain an explicit formula for tautological integrals over these boundary loci. This produces a Gathmann recursion formula for relative invariants in genus one.

@@

By pulling back the equation defining $H$ along $f$ and then taking its $\alpha_k$-th derivative at $x_k$ (which makes sense since all the lower-order derivatives vanish by assumption) we obtain a section of a jet bundle on $\VZ$ whose vanishing locus coincides (set-theoretically) with the locus where $x_k$ belongs to an internal component. As in \cite[Construction 2.1]{Ga} there is an exact sequence of jet bundles
\begin{equation*} 0\to x_k^*\omega_{\mathcal{C}/\VZ}^{\otimes\alpha_k}\otimes\ev_k^*\OO_{\PP^N}(H)\to x_k^*\mathcal J^{\alpha_k}(f^*\OO_{\PP^N}(H))\to x_k^*\mathcal J^{\alpha_k-1}(f^*\OO_{\PP^N}(H))\to 0 \end{equation*}
and thus we obtain a section of the line bundle $x_k^*\omega_{\mathcal{C}/\VZ}^{\otimes\alpha_k}\otimes\ev_k^*\OO_{\PP^N}(H)$ with vanishing locus equal to the locus where $x_k$ belongs to an internal component. Writing this as
\begin{equation*} \Dcal^k = \Dcal_{1,\alpha}^k(\PP^N|H,d) \subseteq \VZ_{1,\alpha}(\PP^N|H,d) = \VZ \end{equation*}
we obtain:
\begin{lem} $(\alpha_k\psi_k+\ev_k^*H)\cap[\VZ]=[\Dcal^k].\footnote{(Navid) Should we work out the vanishing orders of the section at this point?}$\end{lem}
\noindent The remainder of this section will be devoted to providing an explicit recursive description of $[\Dcal^k]$ in terms of tautological classes on spaces of maps with smaller numerical invariants. As we will see, this provides an algorithm for computing reduced relative invariants, as well as a quantum Lefschetz algorithm for reduced (absolute) invariants in genus one.

\subsection{Recursive description of the divisors: types $A,B$ and $C^+$}
We begin by noting that every irreducible component of $\Dcal^k$ is a codimension--$1$ logarithmic stratum. This is obvious: the locus where the log structure is trivial coincides precisely with the nice locus, where the source curve is smooth and not mapped inside $H$. By definition $\Dcal^k$ is disjoint from this locus, and since it has codmension $1$ it must be equal to a union of logarithmic strata.

As discussed at length in the previous section, the divisorial logarithmic strata of $\VZ$ correspond to the rays of the polyhedral subdivision of the tropical moduli space which produces the Artin fan $\Acal$. The logarithmic strata contained in $\Dcal^k$ correspond precisely to the rays of the tropical moduli space where the vertex of the tropicalisation containing $x_k$ is mapped into the interior of $\RR_{\geq 0}$ by the tropical map.

We proceed as in the previous section, fixing an element of $\Mcal$ and considering the resulting subdivision of the tropical moduli space. In fact, it makes sense to choose a locally closed substack of $\Mcal$ where the combinatorial type of the tropical curve (i.e. the ghost sheaf) is constant. Then as long as the minimal subcurve of genus one is not contracted, the picture is classical (i.e. the subdivision is trivial, and $\VZ=\Mcal$ locally). In this case there can only be one vertex of the tropical curve mapped into the interior of $\RR_{\geq 0}$; if there were more, the positions of their images in $\RR_{\geq 0}$ would be independent and thus the corresponding maximal cone in the tropical moduli space would be at least $2$~dimensional, producing a locus in $\VZ$ of larger codimension.\footnote{(Luca) Say something about rigid tropical curves as in ACGS and KLR} We adopt the terminology of Vakil, distinguishing three cases as below:

\begin{figure}[h]
    \centering
    \begin{minipage}{0.3\textwidth}
        \centering
        \Yagraph
        \caption{$\mathcal Y_A$}
    \end{minipage}\hfill
    \begin{minipage}{0.3\textwidth}
        \centering
        \Ybgraph
        \caption{$\mathcal Y_B$}
    \end{minipage}\hfill
    \begin{minipage}{0.3\textwidth}
        \centering
        \Ycgraph
        \caption{$\Ycal_C^+$}
    \end{minipage}
\end{figure}
\noindent Notice that the continuity of the tropical map identifies all of the edge lengths, up to weights given by the expansion factors (i.e. tangency orders) at the edges; thus the tropical moduli space is isomorphic to $\RR_{\geq 0}$ and so we have a divisor (with no subdivision required).

These divisors live inside an open neighbourhood of the preimage in $\VZ$ of our chosen locus in $\Mcal$, and are thus locally closed substacks of $\VZ$. To obtain the desired components of $\Dcal^k$, we must take the closure of these locally closed substacks. We will analyse this procedure on a case-by-case basis, and see once again that the tropical picture gives a complete and recursive description of the combinatorics. As usual, this will be done by comparing loci in$\VZ$ with loci in $\Mcal$. To be more precise: we will take the closure of our locally closed substack in $\Mcal$ and relate it to the closure of our locally closed substack in $\VZ$. We will see that the induced map on closures is another blow-up. The closed locus in $\Mcal$ is understood by \cite{PuncturedMaps} as a fibre product of moduli spaces of punctured maps. \footnote{(Navid) Introduce notation for these locally closed substacks, because this shit is getting out of hand.}

>>>>

\noindent  In the previous section we explained how to determine the logarithmic strata of $\VZ$ along with their dual intersection complex. We now focus our attention on certain logarithmic divisors, providing a recursive description for these loci: that is, a description in terms of fibre products of smaller moduli spaces. This is the technically most difficult part of the paper. The recursive description bears some similarities with the recursive desription of the boundary of the absolute space $\VZ_{1,n}(\PP^N,d)$ given in \cite{VZ}; in keeping with the theme of this paper, however, we will see that log geometry provides a convenient framework for dealing with the combinatorics.

The end result is that we obtain an explicit formula for tautological integrals over these boundary loci. This produces a Gathmann recursion  for relative invariants in genus one.

@@

By pulling back the equation defining $H$ along $f$ and then taking its $\alpha_k$-th derivative at $x_k$ (which makes sense since all the lower-order derivatives vanish by assumption) we obtain a section of a jet bundle on $\VZ$ whose vanishing locus coincides (set-theoretically) with the locus where $x_k$ belongs to an internal component. As in \cite[Construction 2.1]{Ga} there is an exact sequence of jet bundles
\begin{equation*} 0\to x_k^*\omega_{\mathcal{C}/\VZ}^{\otimes\alpha_k}\otimes\ev_k^*\OO_{\PP^N}(H)\to x_k^*\mathcal J^{\alpha_k}(f^*\OO_{\PP^N}(H))\to x_k^*\mathcal J^{\alpha_k-1}(f^*\OO_{\PP^N}(H))\to 0 \end{equation*}
and thus we obtain a section of the line bundle $x_k^*\omega_{\mathcal{C}/\VZ}^{\otimes\alpha_k}\otimes\ev_k^*\OO_{\PP^N}(H)$ with vanishing locus equal to the locus where $x_k$ belongs to an internal component. Writing this locus as
\begin{equation*} \Dcal^k = \Dcal_{1,\alpha}^k(\PP^N|H,d) \subseteq \VZ_{1,\alpha}(\PP^N|H,d) = \VZ \end{equation*}
we obtain:
\begin{lem} $(\alpha_k\psi_k+\ev_k^*H)\cap[\VZ]=[\Dcal^k].\footnote{(Navid) Should we work out the vanishing orders of the section at this point?}$\end{lem}
\noindent The remainder of this section will be devoted to providing an explicit recursive description of $[\Dcal^k]$ in terms of tautological classes on spaces of maps with smaller numerical invariants. As we will see, this provides an algorithm for computing reduced relative invariants, as well as a quantum Lefschetz algorithm for reduced (absolute) invariants in genus one.

\subsection{Recursive description of the divisors: types $A,B$ and $C^+$}
We begin by noting that every irreducible component of $\Dcal^k$ is a codimension--$1$ logarithmic stratum. This is not hard to see: the locus where the log structure is trivial coincides precisely with the nice locus, where the source curve is smooth and not mapped inside $H$. By definition $\Dcal^k$ is disjoint from this locus, and since it has codmension $1$ it must be equal to a union of logarithmic strata.

In the previous section we discussed at length a procedure for determining all the logarithmic strata of $\VZ$. In particular, each divisorial logarithmic stratum is obtained by first choosing the combinatorial type of a tropical map, then subdividing its tropical moduli space, and finally choosing a ray in this subdivision. The logarithmic divisors which are contained in $\Dcal^k$ correspond precisely to the rays of the tropical moduli space where the vertex of the tropicalisation containing $x_k$ is mapped into the interior of $\RR_{\geq 0}$ by the tropical map.

Let us therefore choose a combinatorial type of a tropical map, corresponding to a logarithmic stratum $\Scal$. As long as the minimal subcurve of genus one has positive degree, the subdivision is trivial since there is no alignment necessary. In this case choose a ray of the tropical moduli space $\Q^\vee_{\RR}$. This corresponds to choosing a logarithmic divisor $\Dcal$ in $\Mcal$ containing our original logarithmic stratum $\Scal$ -- the corresponding divisor in $\VZ$ will then be the proper transform of $\Dcal$. In order to have $\Dcal \subseteq \Dcal^k$ we need that the vertex of the tropicalisation containing $x_k$ is mapped into the interior of $\RR_{\geq 0}$ by the tropical map. We claim that this can be the only vertex of the tropical curve mapped into the interior of $\RR_{\geq 0}$; if there were more, the positions of their images in $\RR_{\geq 0}$ would be independent and thus the corresponding maximal cone in the tropical moduli space would be at least $2$~dimensional, which is a contradiction since we assumed we had chosen a ray.\footnote{(Luca) Say something about rigid tropical curves as in ACGS and KLR} There are three cases to distinguish, depending on the position of the circuit. Adopting the terminology of Vakil, we write these as:

\begin{figure}[h]
    \centering
    \begin{minipage}{0.3\textwidth}
        \centering
        \Yagraph
        \caption{$\mathcal Y_A$}
    \end{minipage}\hfill
    \begin{minipage}{0.3\textwidth}
        \centering
        \Ybgraph
        \caption{$\mathcal Y_B$}
    \end{minipage}\hfill
    \begin{minipage}{0.3\textwidth}
        \centering
        \Ycgraph
        \caption{$\Ycal_C^+$}
    \end{minipage}
\end{figure}
\noindent (These are just cartoon pictures: we have omitted the marked points and the expansion factors.) Notice that the continuity of the tropical map identifies all of the edge lengths, up to weights given by the expansion factors (i.e. tangency orders) at the edges; thus the tropical moduli space is isomorphic to $\RR_{\geq 0}$ and so we have a divisor as expected.

Let $\Dcal_{\Mcal}^\circ\subseteq \Mcal$ denote the locally closed stratum corresponding to one of these combinatorial types. The corresponding divisor in $\Mcal$ is $\Dcal_\Mcal = \overline{\Dcal^\circ}$. On the other hand, the corresponding divisor $\Dcal$ in $\VZ$ is the proper transform of $\Dcal_\Mcal$, which is obtained by taking the preimage $\Dcal^\circ$ of $\Dcal_\Mcal^\circ$ (since $\Dcal^\circ$ is disjoint from the blown-up locus) and then taking its closure in $\VZ$. This is different from taking the preimage of $\Dcal_\Mcal$ (the total transform), since in general $\Dcal_\Mcal$ will intersect the blown-up locus. We have an induced map
\begin{equation*} \Dcal \to \Dcal_\Mcal \end{equation*}
which we will see can also be interpreted as a blow-up. This is useful, because $\Dcal_\Mcal$ has a recursive description in terms of moduli spaces of punctured maps \cite{PuncturedMaps}, which in this case we are able to integrate over.
